Emulating Comparative Oncology Trials with Real-world Evidence Studies (ENCORE): Process Development and Methodological Considerations for Oncology Real-World Data

### 2025-07-08 manuscript/manuscript.tex
Manuscript updated

\emph{Manuscript last updated: 2025-07-06 16:21:00.006613}

@@

However, the comparability and transportability of results derived
between observational studies and RCTs depends on multiple factors.
Frequently referenced limitations include lack of baseline randomization
and the imbalance in prognostic factors resulting thereof, missing data,
unmeasured confounding, small study sizes, data
discontinuity,\textsuperscript{3,4} adoption of changes in treatment
guidelines over time in routine clinical care and the inability to
measure and emulate common eligibility criteria, including prognostic
factors, and standardized response assessments in real-world data
(RWD).\textsuperscript{5} While examples of oncology trial emulations
have been published,\textsuperscript{5--7} a systematic and scaled
approach to emulate a diverse set of oncology trials with multiple
heterogeneous databases is necessary to increase confidence in the
interpretability of observational studies, evaluate regulatory
considerations, and to provide context as to which questions can be
validly answered.

The RCT DUPLICATE initiative\textsuperscript{8} increased our
understanding of when observational studies can come to causal
conclusions on treatment effects by benchmarking results against RCTs
under the assumption that each RCT finding reflects a causal treatment
effect. In settings where the RCT designs could be emulated well,
observational studies were able to reach similar
conclusions.\textsuperscript{9} However, this prior work from
RCT-DUPLICATE focused primarily on emulating trials in non-oncology
settings using claims databases.

@@

The focus of ENCORE is to efficiently explore in what contexts
observational studies can or cannot yield similar results compared to
oncology RCTs. Therefore, the project emphasizes trials of therapies for
common cancers for which there has been substantial therapeutic
development in recent years. After review and collaboration with
clinical and regulatory experts, four cancer indications were
identified: non-small cell lung cancer, breast cancer, colorectal cancer
and multiple myeloma. For each cancer type, we aim to conduct three
trial emulations using multiple accessible databases (i.e., the total
equaling 12 trial emulations x \emph{n} databases which are found
fit-for-purpose for each trial).

@@

Causal analyses of observational data are often not designed with
respect to formal hypothesis testing and statistical power in the same
manner as RCTs since the number of `recruited' patients is limited by
available data .\textsuperscript{28} For this project, however, the
emulations in EHR data must have at least equal power to the relevant
trial for interpretation of the pre-specified agreement metrics between
RCT and RWD results. Since the main outcome of interest is defined as
time from index treatment initiation to all-cause mortality (rwOS), the
estimation of the statistical power is driven by the number of events
rather than the number of patients. To assess if the overall number of
events, unstratified by exposure, is sufficient such that a significant
difference can be detected based on the original RCT-reported hazard
ratio (HR), the statistical power (1-\(\beta\)) will be estimated using
Schoenfeld's sample-size formula for the proportional-hazards regression
model.\textsuperscript{29}

>>>>

\emph{Manuscript last updated: 2025-07-08 18:15:17.700176}

@@

However, the comparability and transportability of results derived
between non-interventional studies and RCTs depends on multiple factors.
Frequently referenced limitations include lack of baseline randomization
and the imbalance in prognostic factors resulting thereof, missing data,
unmeasured confounding, small study sizes, data
discontinuity,\textsuperscript{3,4} adoption of changes in treatment
guidelines over time in routine clinical care and the inability to
measure and emulate common eligibility criteria, including prognostic
factors, and standardized response assessments in real-world data
(RWD).\textsuperscript{5} While examples of oncology trial emulations
have been published,\textsuperscript{5--7} a systematic and scaled
approach to emulate a diverse set of oncology trials with multiple
heterogeneous databases is necessary to increase confidence in the
interpretability of non-interventional studies, evaluate regulatory
considerations, and to provide context as to which questions can be
validly answered.

The RCT DUPLICATE initiative\textsuperscript{8} increased our
understanding of when non-interventional studies can come to causal
conclusions on treatment effects by benchmarking results against RCTs
under the assumption that each RCT finding reflects a causal treatment
effect. In settings where the RCT designs could be emulated well,
non-interventional studies were able to reach similar
conclusions.\textsuperscript{9} However, this prior work from
RCT-DUPLICATE focused primarily on emulating trials in non-oncology
settings using claims databases.

@@

The focus of ENCORE is to efficiently explore in what contexts
non-interventional studies can or cannot yield similar results compared
to oncology RCTs. Therefore, the project emphasizes trials of therapies
for common cancers for which there has been substantial therapeutic
development in recent years. After review and collaboration with
clinical and regulatory experts, four cancer indications were
identified: non-small cell lung cancer, breast cancer, colorectal cancer
and multiple myeloma. For each cancer type, we aim to conduct three
trial emulations using multiple accessible databases (i.e., the total
equaling 12 trial emulations x \emph{n} databases which are found
fit-for-purpose for each trial).

@@

Causal analyses of non-interventional data are often not designed with
respect to formal hypothesis testing and statistical power in the same
manner as RCTs since the number of `recruited' patients is limited by
available data .\textsuperscript{28} For this project, however, the
emulations in EHR data must have at least equal power to the relevant
trial for interpretation of the pre-specified agreement metrics between
RCT and RWD results. Since the main outcome of interest is defined as
time from index treatment initiation to all-cause mortality (rwOS), the
estimation of the statistical power is driven by the number of events
rather than the number of patients. To assess if the overall number of
events, unstratified by exposure, is sufficient such that a significant
difference can be detected based on the original RCT-reported hazard
ratio (HR), the statistical power (1-\(\beta\)) will be estimated using
Schoenfeld's sample-size formula for the proportional-hazards regression
model.\textsuperscript{29}
